---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-"
)
```

# humangentools

<!-- badges: start -->
[![pkgdown](https://img.shields.io/badge/docs-pkgdown-blue.svg)](https://jasonahodgson.github.io/humangentools/)
<!-- badges: end -->

humangentools provides sample, population, and SNP ascertainment panel
metadata for common human genomics reference datasets -- currently HGDP
and 1000 Genomes -- to support [tidypopgen](https://cran.r-project.org/package=tidypopgen)
analyses. It's designed to work alongside
[dplaceR](https://jasonahodgson.github.io/dplaceR/) for cross-cultural
context on the same populations.

## Installation

You can install the development version of humangentools from
[GitHub](https://github.com/) with:

``` r
# install.packages("remotes")
remotes::install_github("jasonahodgson/humangentools")
```

## Example

Look up population and region information for a set of sample IDs, and
population-level metadata for a region or dataset:

```{r example}
library(humangentools)

get_sample_information(c("HGDP00001", "HGDP00003"))

get_pop_info(dataset = "HGDP", region = "Central_Asia", exclude = "reference")
```

Browse the Axiom Human Origins array's ascertainment panels, and get the
SNP RS ids for one:

```{r}
get_panel_information()

head(get_axiom_snps("panel1"))
```

See `vignette("humangentools")` for more.
