---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-"
)
```

# humangentools

<!-- badges: start -->
[![pkgdown](https://img.shields.io/badge/docs-pkgdown-blue.svg)](https://jasonahodgson.github.io/humangentools/)
<!-- badges: end -->

humangentools provides sample, population and SNP ascertainment panel metadata for
widely used human genomics reference datasets, to support
[tidypopgen](https://cran.r-project.org/package=tidypopgen) analyses. It is built
to work alongside [dplaceR](https://jasonahodgson.github.io/dplaceR/) for
cross-cultural context on the same populations.

Datasets covered: HGDP, 1000 Genomes, SGDP, HapMap, the AADR (ancient and
present-day), and several individual-study collections.

## Installation

```r
# install.packages("remotes")
remotes::install_github("jasonahodgson/humangentools")
```

## The canonical population code

The same group is often genotyped by several projects, and those genotypes are not
interchangeable — different platforms, different ascertainment. So each
(group, dataset) pair gets its own `pop` code:

```{r example}
library(humangentools)

get_pop_info(population = "Yoruba")[, c("pop", "source_dataset", "n_samples")]
```

Sample lookup returns one row per (sample, dataset), so pass `dataset` when you
need exactly one row per sample:

```{r}
get_sample_information("HGDP00001", dataset = "HGDP")
```

Populations are classified as present-day or ancient, and filterable:

```{r}
table(get_pop_info()$temporal)
```

## SNP ascertainment panels

```{r}
get_panel_information()

length(get_axiom_snps("panel4"))
```

## Linking to D-PLACE societies

```{r}
get_dplace_link(population = "Yoruba")[, c("pop", "soc_id", "society_name",
                                           "n_marriage_vars_coded")]
```

See `vignette("humangentools")` for more.
