---
title: "Getting started with humangentools"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with humangentools}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(humangentools)
```

humangentools bundles sample, population and SNP ascertainment panel metadata for
widely used human genomics reference datasets, so that
[tidypopgen](https://cran.r-project.org/package=tidypopgen) analyses can be
annotated and filtered without tracking that metadata down each time. It is built
to work alongside [dplaceR](https://jasonahodgson.github.io/dplaceR/) for
cross-cultural context on the same populations.

## Populations are keyed by a canonical code

The central idea is the `pop` code. The same human group is often genotyped by
several projects, and those genotypes are **not interchangeable**: they come from
different arrays or sequencing platforms, with different ascertainment and
different error profiles. So each (group, dataset) pair gets its own code,
following the convention of `kgp::allmeta` — a bare three-letter code for 1000
Genomes populations, and `<Name><DATASET>` elsewhere.

```{r}
get_pop_info(population = "Yoruba")[, c("pop", "source_dataset", "n_samples")]
```

Four codes, four genotype datasets, one ethnic group. Pooling them under a single
label would mix array and sequence data, and any resulting difference in
heterozygosity or call rate would look like biology.

The datasets available:

```{r}
get_dataset_info()[, c("dataset", "n_pops", "n_samples", "genotyping")]
```

## Sample metadata

`get_sample_information()` looks up the canonical code, population label, region
and source dataset for one or more sample IDs:

```{r}
get_sample_information("HGDP00001")
```

Two rows for one sample, because the AADR regenotyped this individual on the
1240K panel. `id` identifies the person and joins across datasets; `source_id` is
the contributing dataset's own key; `data_type` records the AADR library type.
Rows are unique on `(source_id, pop)`, not on `(id, pop)`.

**If you are attaching metadata to a `gen_tibble`, pass `dataset`** — otherwise
you get more rows than samples and any column-wise assignment silently
misaligns:

```{r}
get_sample_information("HGDP00001", dataset = "HGDP")
```

An ID that is not found is kept with `NA` fields and a warning; `na.fill = FALSE`
drops it instead:

```{r}
get_sample_information(c("HGDP00001", "not-a-real-id"), na.fill = FALSE)
```

## Population metadata

`get_pop_info()` goes the other way: given a dataset, region, population or set of
sample IDs (including a `gen_tibble`), it returns population-level metadata.

```{r}
get_pop_info(dataset = "HGDP", region = "Central_Asia",
             include = c("pop", "population_label", "temporal", "lat", "lon"))
```

### Ancient and present-day populations

The AADR contributes several thousand **ancient** populations alongside
present-day ones. Anything joined to ethnographic data wants present-day
populations only, so `temporal` is a filter:

```{r}
table(get_pop_info()$temporal)
```

```{r}
get_pop_info(temporal = "ancient", dataset = "AADR",
             include = c("pop", "population_desc", "lat", "lon"))[1:5, ]
```

### Two sets of coordinates

Where a group was sampled is not always where it comes from. Both are stored, and
`location` chooses which pair lands in the convenience columns `lat` and `lon`:

```{r}
get_pop_info(pop = "GIH", location = "origin")[, c("pop", "lat", "lon")]
get_pop_info(pop = "GIH", location = "sampling")[, c("pop", "lat", "lon")]
```

GIH is Gujarati Indians sampled in Houston. For some populations an origin is not
a point at all — African Caribbean in Barbados originates in a distribution over
West and Central African source populations — and those carry `NA` origin
coordinates with the reason in `coord_note`:

```{r}
get_pop_info(pop = c("ACB", "ASW"))[, c("pop", "origin_lat", "coord_note")]
```

## SNP ascertainment panels

The Axiom Human Origins array was ascertained from thirteen distinct panels, each
discovered by sequencing one individual. Which panel a SNP came from governs how
its allele frequencies behave across populations, so it matters for any
comparison of diversity between groups.

```{r}
get_panel_information()
```

`get_axiom_snps()` returns the RS ids for one or more panels:

```{r}
length(get_axiom_snps("panel4"))
```

Panels are **not** mutually exclusive. Panel 4 is the San panel, ascertained as
heterozygous sites in HGDP01029, but only about 43% of its SNPs are unique to it:
the rest were also heterozygous in other ascertainment individuals, which is a
different and frequency-biased discovery process. For a single ascertainment
history, subtract every other panel:

```{r}
san <- setdiff(get_axiom_snps("panel4"), get_axiom_snps(paste0("panel", c(1:3, 5:13))))
c(panel4 = length(get_axiom_snps("panel4")), exclusive = length(san))
```

Note that panel 13 is subtracted along with the others. It is labelled
`Denisova-San` and its ascertainment sample is `Den-HGDP01029`, so although the
San genome is involved, the discovery criterion contrasts it with Denisova — a
second process rather than more of the same one. Check
`get_panel_information()` before assuming two panels share an ascertainment
history.

## Linking to dplaceR societies

`get_dplace_link()` returns the crosswalk joining these populations to societies
in [dplaceR](https://jasonahodgson.github.io/dplaceR/)'s D-PLACE data, on a
common `soc_id`:

```{r}
get_dplace_link(population = "Yoruba")[, c("pop", "soc_id", "society_name",
                                           "confidence", "n_marriage_vars_coded")]
```

Every code carrying the label gets the same society, because a society link is a
property of the population and not of the sequencing project.

The table has been reviewed by hand: `reviewed` is `TRUE` on confirmed rows, and
rejected candidates have been removed rather than left to be re-examined. A
population with no society still appears, with `soc_id = NA`, so the crosswalk
records a checked dead end rather than an absence — `include_unmatched = TRUE`
shows those.

`n_marriage_vars_coded` counts how many of ten Ethnographic Atlas marriage and
descent variables (EA009, EA012, EA015, EA018, EA020, EA023–EA026, EA043) have a
genuine coded observation for the linked society, missing-data sentinels
excluded. A link can be ethnographically correct and still analytically useless
if the society has no coded variables, so filter on it:

```{r}
usable <- get_dplace_link(min_marriage_vars = 8)
c(rows = nrow(usable),
  populations = length(unique(usable$pop)),
  societies = length(unique(usable$soc_id)))
```

That count, rather than any SNP total, is usually what limits a study combining
these resources.
