---
title: "Getting started with humangentools"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with humangentools}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(humangentools)
```

humangentools bundles sample, population, and SNP ascertainment panel
metadata for common human genomics reference datasets (currently HGDP and
1000 Genomes), so that [tidypopgen](https://cran.r-project.org/package=tidypopgen)
analyses of these datasets can be annotated and filtered without tracking
down that metadata yourself each time. It's designed to work alongside
[dplaceR](https://jasonahodgson.github.io/dplaceR/) for cross-cultural
context on the same populations.

## Sample and population metadata

`get_sample_information()` looks up the population and region for one or
more sample IDs:

```{r}
get_sample_information(c("HGDP00001", "HGDP00003"))
```

By default, an ID that isn't found is kept in the result with `NA`
population/region (and a warning); set `na.fill = FALSE` to drop it
instead:

```{r}
get_sample_information(c("HGDP00001", "not-a-real-id"), na.fill = FALSE)
```

`get_pop_info()` goes the other way: given a dataset, region, or population
(or a set of sample IDs, including a `tidypopgen` `gen_tibble`), it returns
population-level metadata -- coordinates, source dataset, and citation:

```{r}
get_pop_info(region = "Central_Asia")
```

```{r}
get_pop_info(samples = c("HGDP00001", "HGDP00003"))
```

## SNP ascertainment panels

The Axiom Human Origins array was ascertained from several distinct
population panels, which matters for some population-genetic analyses.
`get_panel_information()` describes the panels available:

```{r}
get_panel_information()
```

`get_axiom_snps()` returns the RS ids for one or more panels, ready to use
as a SNP filter (e.g. before reading genotype data into a `tidypopgen`
`gen_tibble`):

```{r}
length(get_axiom_snps("panel1"))
length(get_axiom_snps("all"))
```
